---
title: "Data Merging and Prediction"
subtitle: "Application: Predicting calorie consumption (NHANES)"
author: "??"
date: "5/14/2024"
output: html_document
---

# 1. Overview

The goal of this analysis is to explore the relationship between total caloric intake, including beer consumption, and demographic factors. We will utilize some important R tools, including data importing, merging, summarizing by groups, linear and log-linear regression, and making predictions.

# 2. Data Summary
 
We will use 2 datasets from the National Health and Nutrition Examination Survey (NHANES). The dataset DEMO includes respondents from 2015-16 and 2017-18. The dataset CALORIES was created using the NHANES datasets DR1IFF_I.XPT and DR1IFF_J.XPT (on daily food consumption) and DRXFCD_J.XPT (on food codes) available at: https://wwwn.cdc.gov/nchs/nhanes/Default.aspx.

**DEMO**

* **SEQN:**          Respondent's sequence number
* **AGE:**           Respondent's age (years)
* **GENDER:**        Respondent's biological sex (Male, Female)
* **EDUCATION:**     Respondent's education level (Less than High School, High School, Some College, College Grad)
* **MARITAL:**       Respondent's marital status (Married, Widowed, Divorced, Separated, Single, Living Partner)
* **PREGNANT:**      Respondent's pregnancy status (Yes, No)

**CALORIES**

* **SEQN:**          Respondent's sequence number
* **ALL_CALORIES:**  Total calories consumed (food + drinks) by respondent in past 24 hours     
* **BEER_CALORIES:** Calories consumed from beer in past 24 hours

Let's first load some useful libraries. 
```{r, message = FALSE}
library(sjPlot)
library(tidyverse)
library(formattable)
library(jtools)
library(scales)
library(ggridges)
library(viridis)
```

Use read.csv() to load the datasets.
```{r}
setwd(??)
DEMO = read.csv("NHANES/DEMO.csv")
CALORIES = read.csv("NHANES/CALORIES.csv")
```

# 3. Data Merging

**QUESTION 1:** Use full_join() to merge the datasets using "SEQN" as the common link. Name this new dataset FULL. How many observations are in FULL?
```{r}
??
```

**QUESTION 2:** Use inner_join() to merge the datasets using "SEQN" as the common link. Name this new dataset INNER. How many observations are in INNER?
```{r}
??
```

We will use INNER in the remainder of the lab, as we want only to include only those individuals with both demographic information and calorie consumption data.

# 4. Data Exploration

**QUESTION 3:** Calculate the mean of ALL calories and BEER calories, by gender. 
```{r}
SUMMARY_GENDER = ??

formattable(SUMMARY_GENDER, digits=3)
```

**QUESTION 4:** Consumers with what education level consume the most TOTAL calories? And the most BEER calories?
```{r}
INNER$EDUCATION = factor(INNER$EDUCATION, levels=c("College Grad", "Some College", "High School", "Less than High School"))

SUMMARY_EDUCATION = ??

formattable(SUMMARY_EDUCATION, digits=3)
```

# 5. Data Visualization

**QUESTION 5:** Create a ridge plot of total calorie consumption, by gender and education level.
```{r message=FALSE, warning = FALSE}
SUMMARY_BOTH = INNER %>%
  group_by(EDUCATION, GENDER) %>%
  summarize(MEAN_ALL=mean(ALL_CALORIES))

ggplot(data=INNER, aes(x=??, y=??, fill=EDUCATION)) + facet_wrap(~GENDER) +
  geom_density_ridges(alpha=0.5) + 
  scale_fill_viridis("Education", discrete = TRUE) +
  scale_x_continuous("Total calories", limits=c(0,6000)) +
  scale_y_discrete("") +
  geom_text(data=SUMMARY_BOTH, aes(x=MEAN_ALL, y=EDUCATION, label=round(MEAN_ALL, digits=0)), nudge_y=0.8) +
  ggtitle("Daily calories consumed, by gender and education")

```

**QUESTION 6:** Create a ridge plot of total calorie consumption, by gender and marital status.
```{r message=FALSE, warning = FALSE}
INNER$MARITAL = factor(INNER$MARITAL, levels=c("Married", "Living Partner", "Divorced", "Separated", "Widowed", "Single"))

SUMMARY_MARITAL = INNER %>%
  group_by(MARITAL, GENDER) %>%
  summarize(MEAN_ALL=mean(ALL_CALORIES))

ggplot(data=INNER, aes(x=??, y=??, fill=MARITAL)) + facet_wrap(~GENDER) +
  geom_density_ridges(alpha=0.5) + 
  scale_fill_viridis("Status", discrete = TRUE) +
  scale_x_continuous("Total calories", limits=c(0,6000)) +
  scale_y_discrete("") +
  geom_text(data=SUMMARY_MARITAL, aes(x=MEAN_ALL, y=MARITAL, label=round(MEAN_ALL, digits=0)), nudge_y=0.8) +
  ggtitle("Daily calories consumed, by gender and marital status")
```

# 6. Linear Regression Analysis

**QUESTION 7:** Regress ALL calories on age, gender, education and marital status using the INNER data. Which variable(s) are significant?
```{r}
regression1 = ??
summ(regression1, digits=2)
```

**QUESTION 8:** Regress ALL calories on gender and pregnancy status using the INNER data. What is your interpretation of the pregnant variable?
```{r}
regression2 = ??
summ(regression2, digits=2)
```

# 7. Logistic Regression Analysis

**QUESTION 9:** Create a new binary variable called BEER_DRINKER if a person consumes any beer calories. How many beer drinkers are in the data?
```{r}
INNER$BEER_DRINKER = ??
table(??)
```

**QUESTION 10:** Run a logistic regression on BEER_DRINKER with age, gender, education and marital status as independent variables. Which variable(s) are significant?
```{r}
logit1 = ??
summ(logit1, exp=TRUE, digits=2)
```

```{r}
plot_model(logit1, show.values = TRUE, value.size=3, value.offset=0.5) +
  ggtitle("Odds ratios for beer drinker") 
```

**QUESTION 11:** Run a logistic regression on BEER_DRINKER with gender and pregnancy status as independent variables. Which variable(s) are significant?
```{r}
logit2 = ??
summ(logit2, exp=TRUE, digits=2)
```

Finally, let's use regression 1 to make a prediction for a friend, a 25-year old single man who graduated college. We first create a small dataframe called FRIEND and enter some input values for the regression.
```{r}
FRIEND = data.frame(
  AGE = 25,
  GENDER = "Male",
  EDUCATION = "College Grad",
  MARITAL = "Single"
)
```

Then we enter the input data into the predict() command. We can also obtain a 95% prediction interval.
```{r}
predict(regression1, newdata = FRIEND, interval = "prediction")
```

**QUESTION 12:** Your neighbor is a 32-year old female Instagram influencer who attended some college and is living with a partner. What is her predicted calorie consumption?
```{r}
??
```